### 2024-11-07 docs/paper.tex
Half-done with Jo's edits

\maketitle
\begin{abstract}
{\begin{spacing}{1.2}
Housing is the largest asset in U.S. household portfolios, and a first-time homebuyers increasingly rely on parental transfers. This paper quantifies the contribution of parental transfers to the homeownership rate of young households. I build and estimate a life-cycle overlapping generations model with housing, where adult children and parents interact without commitment. I find that parental transfers account for 14 percentage points (29\%) of young households' homeownership. Transfers from wealthy parents not only help households overcome borrowing constraints but also help them maintain homeownership, reducing the downsides of illiquidity. Policies lowering entry barriers to homeownership generally increase the reliance on parental wealth, while increasing liquidity lowers the role of parental wealth. Finally, I study how illiquidity affects the commitment problem  between parents and children. Wealthy parents' children increase investment in illiquid housing to extract larger transfers and exhibit a preference for illiquidity.
\end{spacing}}
\end{abstract}
\newpage
\onehalfspacing

\section{Introduction}
Housing is the largest asset in U.S. households' portfolios and homeownership is seen as central to the ``American Dream'' \citep{goodman2018homeownership}. Yet access to homeownership appears to be increasingly reliant on parental transfers. Around 30\% of American first-time homebuyers received downpayment assistance from their parents between 2009 and 2016. Parental downpayment assistance averaged \$48,000 among receiving households.\footnote{Data from the Survey of Household Economic Decisionmaking and the Transfer Supplement of the 2013  Panel Study of Income Dynamics (see Section \ref{sec:data}).}  It is important to understand how transfers shape households' ability to access and maintain ownership, given  housing’s importance to household portfolios, economic outcomes \citep{sodini2023identifying}, and business cycles \citep{Kaplan2020,favilukis2017macroeconomic}.

The main objective of this paper is to quantify the importance of parental transfers for young adults' homeownership. The primary mechanism by which transfers increase homeownership, and the focus of most empirical studies, is through relaxing borrowing constraints \citep[see e.g.,][]{Lee2018,Blickle2019,wold2024housing}. A secondary mechanism is that parental transfers help households maintain homeownership, a mechanism that has received less attention  \citep[see e.g.,][]{bond2021role}. However, while these studies show that receiving transfers causally increases homeownership, they do not address the overall importance of parental transfers. For example, households might anticipate future transfers and save less. In this case, even if transfers increase the likelihood of homeownership all else equal, they may not affect aggregate homeownership. The secondary objective is to study how policies lowering homeownership barriers interact with parental transfers.

I first present empirical results suggesting that parental wealth not only helps in the transition to ownership but also in maintaining ownership. I show that, controlling for a wide set of household and parental characteristics, households with wealthier parents not only purchase pricier homes but are also less likely to fall behind on mortgage payments after becoming homeowners. Finally, in a variation of the event study in \cite{Chetty2007}, I show that while most households downsize their housing during unemployment spells, those with wealthy parents retain their current homes.

To answer the research questions, I combine two distinct models: housing models, which ignore parental transfers, and models of parental transfers, which ignore housing. I build a life-cycle overlapping generations model with altruistic parents and a rent-or-own decision. Parents and children  interact without commitment through both inter-vivos transfers and bequests. In models of altruism without housing (e.g., \citealp{Altonji1997a,Barczyk2014}), the transfer motive is to increase the child's consumption when they are borrowing constrained and thus have high marginal utility of wealth. The inclusion of a frictional housing market generates two new transfer motives. First, transfers directly alleviate credit constraints, and so wealthy parents may transfer to allow their children to meet the downpayment requirement. Second, future transfers provide partial insurance against future income shocks, reducing risks associated with the large illiquid investment of homeownership. Moreover, the illiquidity of housing serves as a commitment device enabling children to secure future transfers. As in all transfer models, the expectation of future transfers reduces children’s savings motives, potentially lowering the homeownership rate.

Beyond computational challenges, what insights do we gain by allowing parental transfers in all periods? After all, previous studies have used alternative assumptions, such as a one-time transfer after which the child is independent at age 24 \citep[e.g.,][]{Lee2019} or warm-glow bequests that are distributed to younger households \citep[e.g.,][]{aliprantis2022dynamics}. The key insight of my modeling approach is that transfers are critical not only for entering homeownership---the main focus of empirical literature---but also for maintaining homeownership. Specifically, the possibility of future transfers mitigates the downsides of homeownership for liquidity-constrained households. This turns out to be as important as relaxing borrowing constraints. Moreover, the model replicates the empirical pattern that households with wealthy parents do not downsize during income losses, which would not be possible without allowing transfers in all periods.

I estimate the model by matching data on homeownership, wealth, and transfers from the Panel Survey of Income Dynamics (PSID). To quantify the importance of  transfers, I find the counterfactual homeownership rate without parental transfers. Without parental transfers, the model simplifies to a standard life-cycle model with housing. I find that parental transfers account for 14 percentage points (29\%) of the young homeownership rate. In the benchmark model, the supply of owner-occupied housing is perfectly elastic; however, my results are robust to alternative supply elasticities.

>>>>

\maketitle
\begin{abstract}
{\begin{spacing}{1.2}
Housing is the largest asset in U.S. household portfolios, and a first-time homebuyers increasingly rely on parental transfers. This paper quantifies the contribution of parental transfers to the homeownership rate of young households. I build and estimate a life-cycle overlapping generations model with housing, where adult children and parents interact without commitment. I find that parental transfers account for 14 percentage points (29\%) of young households' homeownership. Transfers from wealthy parents not only help households overcome borrowing constraints but also help them maintain homeownership, reducing the downsides of illiquidity. Policies lowering entry barriers to homeownership generally increase the reliance on parental wealth, while increasing liquidity lowers the role of parental wealth. Finally, I study how illiquidity affects the commitment problem and find that children with wealthy parents use the illiquidity of housing to encourage transfers and exhibit a preference for illiquidity.
\end{spacing}}
\end{abstract}
\newpage
\onehalfspacing

\section{Introduction}
Housing is the largest asset in U.S. households' portfolios. While homeownership is seen as central to the ``American Dream'' \citep{goodman2018homeownership}, households face different barriers to ownership. In particular, around 30\% of American first-time homebuyers received downpayment assistance from their parents between 2009 and 2016. Parental downpayment assistance averaged \$48,000 among receiving households.\footnote{Data from the Survey of Household Economic Decisionmaking and the Transfer Supplement of the 2013  Panel Study of Income Dynamics (see Section \ref{sec:data}).}  It is important to understand how transfers shape households' ability to access and maintain ownership, given  housing’s importance to household portfolios, economic outcomes \citep{sodini2023identifying}, and business cycles \citep{Kaplan2020,favilukis2017macroeconomic}.

This paper quantifies the importance of parental transfers for young adults' homeownership and investigates how policies lowering homeownership barriers interact with parental transfers. Previous empirical studies highlight that parental transfer receipt is associated with an increase in the probability of becoming a homeowner, mainly by relaxing credit constraints.\citep[see e.g.,][]{Lee2018,Blickle2019,wold2024housing}. Additionally. parental transfers help households maintain homeownership, a mechanism that has received less attention \citep[see e.g.,][]{bond2021role}. However, while these studies show that receiving transfers causally increases homeownership, they do not address the overall importance of parental transfers. For example, households may anticipate future transfers and subsequently save less. In this case, even if transfers increase the likelihood of homeownership all else equal, they may not affect aggregate homeownership.

I first present empirical evidence indicating that parental wealth helps not only in the transition to ownership, but also in maintaining ownership. I show that, controlling for a wide set of household and parental characteristics, households with wealthier parents not only purchase pricier homes but are also less likely to fall behind on mortgage payments after becoming homeowners. Finally, using a variation of the event study in \cite{Chetty2007}, I show that while most households downsize their housing during unemployment spells, those with wealthy parents retain their current homes.

To quantify the impact of parental transfers on homeownership, I then combine two distinct models: housing models, which ignore parental transfers, and models of parental transfers, which ignore housing. I build a life-cycle overlapping generations model with altruistic parents and a rent-or-own decision. Parents and children interact without commitment through both inter-vivos transfers and end-of-life bequests. In models of altruism without housing (e.g., \citealp{Altonji1997a,Barczyk2014}), the transfer motive is to increase the child's consumption when they are borrowing constrained, thus having high marginal utility of wealth. The inclusion of a frictional housing market generates two new transfer motives. First, transfers can directly alleviate credit constraints, enabling wealthier parents to help their children meet downpayment requirements. Second, future transfers provide partial insurance against future income shocks, reducing risks associated with the large illiquid investment of homeownership. Moreover, the illiquidity of housing serves as a commitment device enabling children to secure future transfers. As in all transfer models, the expectation of future transfers reduces children’s savings motives, potentially lowering the homeownership rate.

Allowing for parental transfers in all periods highlights that transfers are critical not only for entering homeownership---the main focus of empirical literature---but also for maintaining homeownership. Specifically, the prospect of future transfers mitigates the downsides of homeownership for liquidity-constrained households. Ameliorating illiquidity turns out to be as important as relaxing borrowing constraints. Crucially, the model rationalizes the previously discussed empirical pattern that households with wealthy parents do not downsize during income losses, which would not be possible without allowing transfers in all periods.

I estimate the model by matching data on homeownership, wealth, and transfers from the Panel Survey of Income Dynamics (PSID). To quantify the importance of transfers, I find the counterfactual homeownership rate without parental transfers. Without parental transfers, the model simplifies to a standard life-cycle model with housing. I find that parental transfers account for 14 percentage points (29\%) of the young homeownership rate.
